---
title: "Inferential Stats with R"
output:
  xaringan::moon_reader:
    css: ["style.css", "default"]
    lib_dir: libs
    df_print: paged
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: 16:9
---

```{r child = "setup.Rmd"}
```

```{r eval = TRUE, echo = FALSE}
```

class: center, middle, dk-section-title

# Welcome to the Course!

---


## What you will learn in this course

???

This course will cover seven topics. 
1. First, we're going to learn four main inferential statistics...

5. We'll also learn how to conduct reliability such as Cronbach's alpha, a measure of internal consistency.

6. Then we'll learn how to report statistics in R through learning how to extract values from the summary output, tidying up the results, and using packages to report statistics easily.

7. Lastly, we'll cover how to test assumptions of your parametric tests and what to do if you do not meet the assumptions of normally distributed data. 

--

1. t-tests: independent and dependent

--

2. ANOVA: mainly the one-way ANOVA

--

3. chi-square

--

4. regression

--

5. reliability (internal consistency)

--

6. how to report statistics in R

--

7. how to test parametric assumptions

---

## What you will need for this course

-- 

1. R and RStudio!

--

2. Dataset: [insert dataset here]

--

3. Packages: [insert packages here]
  + broom
  + apa
  + psych
  + lm.beta
  + etc.
  
---

## How this course will work




