---
title: "Creating the dataset"
author: "DLW"
date: "5/20/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(wakefield)
library(tidyverse)
library(summarytools)
library(psych)
library(scales)
library(janitor)
```

```{r}
dat <-
  r_data_frame(
    n = 300,
    id = id,
    # 2 groups
    athlete = dummy,
    smokes = smokes,
    live_on_campus = dummy,
    # 3+ groups
    race = race, 
    sex = sex_inclusive,
    `grade_class` = level,
    political = political,
    # Scale variables
    height = height,
    iq = iq(mean = 103.5),
    hs_gpa = normal(mean = 3.1, sd = .3, min = 2, max = 4),
    r_series(normal, j = 2, name = "weight", mean = 170, sd = 15, relate = "+5_5"),
    `age` = rpois(lambda = 3), #Need to add 17 to this value
    r_series(normal, j = 4, name = "ACT", mean = 21, sd = 5.4, min = 1, max = 36, relate = "+.3_6"),
    r_series(likert, 5, name = "SWLS_time1", relate = "-.5_.1"),
    r_series(likert, 5, name = "SWLS_time2", relate = "-.5_.1"),
    r_series(normal, j = 3, name = "exam", mean = 75, sd = 10, min = 0, max = 100, relate = "+5_5")
    ) %>%
  mutate_at(vars(starts_with("SWLS")), rescale, to = c(1, 5)) %>% # Comment 1 below
  mutate_at(vars(starts_with("SWLS")), round, digits = 0) %>% 
  mutate_at(vars(starts_with("ACT")), rescale, to = c(1, 36)) %>% 
  mutate_at(vars(starts_with("Exam")), rescale, to = c(0, 100)) %>% 
  mutate_at(vars(starts_with("ACT"), starts_with("weight"), starts_with("Exam")), 
            round, digits = 0) %>%
  rename(ACT_English = ACT_4, # Comment 2 below
         ACT_Mathematics = ACT_3,
         ACT_Reading = ACT_1,
         ACT_Science = ACT_2) %>% 
  clean_names()
```

# Comment 1

Unfortunately, the relate argument in the r_series option messes up the range of the response scale. If you ran the above code for SWLS without the mutate_at function below, then this would be the response ranges for the five items:

1. Item_1: 1-5
2. Item_2: 0-5
3. Item_3: -1-5
4. Item_4: -2-5
5. Item_5: -3-4

Clearly, the response scales should always be 1-5. Therefore, needed to rescale using the scales::rescale package into being 1-5. 

# Comment 2

ACT scores range from 1-36 and the average composite score is 21 (sd = 5.4). I did the relate function so that the scores went up by subscale slightly and then renamed and ordered by what students typically get highest scores on (english) to lowest (science). 

```{r}
# So I can quickly see the data summary
print(dfSummary(dat, plain.ascii = FALSE, style = "grid", valid.col = FALSE), method = 'render')
```

```{r}
# Needed to check my scales had decent Cronbach's alphas
dat %>%
  as.data.frame() %>%
  select(starts_with("SWLS_time2")) %>%
  psych::alpha()

# Needed to check my ACT scores had the proper-ish relationships
cor(dat[, c(16:19)])
```

```{r}
write_csv(dat, path = "dataset.csv") 
```


```{r}
temp <- read_csv("dataset.csv")
```

